Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-01-15 manuscript/manuscript.Rmd
General discussion

# General Discussion
Advice substantially shapes people's career and personal outcomes.
Therefore, understanding advisers' motivations has far-reaching
implications. We propose that advisers' focus on advisees' expectations
and the desire to avoid disappointing them can induce gender differences
in the advice given and ultimately in the outcomes achieved. Gender
differences in career outcomes contribute significantly to the gender
wage gap, with women often holding less senior positions and
experiencing worse career outcomes, even with equal performance.
Therefore, identifying a factor such as expectations that contributes to
this gap has important implications for organizational practice.

In this paper, we examine a novel and unexplored factor of
advice-giving: interpersonal considerations. Advisers who take into
account the advisee's belief utility inflate their advice to convey a
more favorable impression, leading to worse decisions and gender
differences. Men have more inflated expectations than women, and taking
expectations into account can lead to gender differences, even without
discrimination. We ruled out discrimination as a separate channel by
withholding information about the gender of the advisee. Finally, we
expect that advisers are aware of the tradeoff between giving accurate
advice and avoiding disappointment and, thus, are less likely to inflate
their advice when advisees can compensate them after observing the
outcome of their choice.

The findings of this study suggest that organizations can promote gender
equity by calibrating employees' expectations or shifting advisers'
attention away from them. While these measures may not entirely solve
the gender gap problem, they can make a meaningful difference when
unconscious bias interventions have failed to show any effect
[@Chang2019; @Paluck2009]. We believe our findings may be particularly
relevant to societies that value signaling in interpersonal
relationships. Future work
could also examine whether people in different cultures place more
weight on expectations.

>>>>

# General Discussion
Advice has the potential to shape people's career and personal outcomes.
Honest feedback, however, may be painful to learn if it falls short of one's
expectations. As prior work notes, this may motivate people to avoid information 
and avoid seeking help [@Benabou2022; @Golman2017; @Jaroszewicz2022].
We present evidence from two experiments that advisers are also cognizant of this
cost. As a result, they present flattering advice that avoids disappointing the
recipient, and correctly anticipate that this boosts how advisees perceive them.

However, this flattering advice comes at a cost to advisees, who would do worse
if they followed it blindly, as we show in Study 1.
<!-- David: But not Study 2? I guess we're not sure yet? -->
Moreover, a desire to avoid disappointment also means that advisers have to
take into account the expectations of the advisee. 
We document that men are more optimistic about their performance than
women are. As a result, they receive more flattering advice and are more likely
to be told to aim higher. Notably, in the context of our experiment, this turns
out to be bad advice ex post.

Our findings have implications for organizational practice, where mentoring
and advice-giving may take into account an employees' expectation. We
document this as a novel source of gender bias. Organizations could reduce
this bias by calibrating employees' expectations to reduce overconfidence.

Participants in our experiment were paired anonymously. Even so, we document
this desire to avoid disappointment. We anticipate that advice would be more
flattering in face-to-face communication and without anonymity. Moreover,
existing relationships might make it even more difficult for advisers to be
honest.

In our experiments, participants only received advice once and did not evaluate
the adviser after observing the outcome of their decision.
For example, advice that leads to bad outcomes may undermine the
interpersonal benefits of flattery. Alternatively, people may still like the
flattering advice and not fault the adviser for the bad outcome.
Moreover, our setting involved only a single piece of advice on one task.
Future research could examine whether people return to those who gave them
flattering advice, or if they prefer someone who gave them the honest
(but unpleasant) truth.

Advice has long been studied from the perspective of the receiver. Similarly,
research on belief utility has examined how recipients respond to the valence
of the information they receive. Here, we show that advisers, too, take into
account the psychological impact of the information they convey. They may have
even greater motivations to avoid conveying bad news, because they incur the
interpersonal costs of delivering unfavorable information without reaping the
benefits from helping someone make a better choice.
